\bf An HMM-based part-of-speech tagger for Turkish

### 2015-12-09 paper/paper.tex
Fix typos

\begin{document}
%include agda.fmt
\maketitle

@@

The main motivation for this project comes from the lack of a POS tagger for
Turkish exclusively implemented in the programming language Haskell. Moreover, I
believe that implementing a POS tagger from scratch is beneficial in terms of
gain experience with relevant statistical models through this process.

Due to time and resource constraints, it is not plausible to build a
state-of-the-art POS tagger for Turkish. Nevertheless, I expect my POS tagger to
work with decent accuracy that is adequate to be a starting point for further
development. Moreover, \citet{Korkut2015} began the development of a Turkish NLP
library named ``Guguk'', implemented in the programming language Haskell. I
intend to merge this project to Guguk, hence contributing towards the existence
of a Haskell Turkish NLP library; I intend to continue development of this
project as a part of Guguk.

Before I proceed, let me give a few remarks on notational conventions I follow
throughout this text: I use \texttt{HMM} (with the fixed-width font) to the
datatype I'm using from the \texttt{Data.HMM} package and HMM (with regular
font) to refer to the concept.

@@

% Briefly survey the most salient prior work that relates to the problem you wish
% to solve. This section should cite relevant sources. This section should be at
% least one to two paragraphs in length.
\cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
This proves an effective approach seeing that (1) Turkish morphology is
exclusively based on suffixes, in contrast to to prefixes (2) morphology is
central to the Turkish language so it is likely that any given word is formed
through the agglutination of at least severals morphemes. The tagger described
by \citet{dincer2008suffix} has ``\%90.2 accuracy in the best case''.

@@

The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which one can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize the \texttt{HMM} datatype since it has three
fields containing functions \footnote{See
  http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.}. I consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development; this has
almost no effect on performance, but it obfuscates the code.

Since processing the corpus is a resource-demanding operation,
\texttt{create-model} is intended be run only once after which \texttt{run-tag}
can be used to make use of the serialized components of the HMM.

I should note that it is not possible to give a complete account of the
program's logic in this chapter although I explain the most salient components.
The source code is in the \texttt{ayberkt-final\_project} repository and it is
well-commented. Should the reader have any concerns regarding implemenetation
they could find it convenient to consult the source code.

@@

The function \texttt{freqMap} found in \texttt{create/Main.hs} takes as input a
list of orderable elements (denoted with \texttt{Ord a}) and creates a
\texttt{Map} containing, as value, the number of ocurrence of each element.

\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ M.Map a Int
freqMap elems = populate empty elems
  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ M.Map a Int
         populate m [] = m
         populate m (x:xs) = let freq = (M.findWithDefault 0 x m) :: Int
                              in populate (M.insert x (freq + 1) m) xs
\end{lstlisting}

This function is called several times with lists of values designate to be
\emph{counted} (such as the list of tag bigrams, or the list of all tagged words)
et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
counts) will need to be accessed many times by the viterbi algorithm when
finding the most probable path through the model. Once the necessary maps are
generated, they are written to files in the directory \texttt{model}.

@@

\texttt{probability} when combined with \texttt{freqMap} forms an efficient
method for implementing dependent probability.

>>>>

\begin{document}
%include agda.fmt

@@

The main motivation for this project comes from the lack of a POS tagger for
Turkish exclusively implemented in the programming language Haskell. Moreover, I
believe that implementing a POS tagger from scratch is beneficial in terms of
gaining experience with relevant statistical models.

Due to time and resource constraints, it is not plausible to build a
state-of-the-art POS tagger for Turkish. Nevertheless, I expect the POS tagger
to work with decent accuracy that is adequate to be a starting point for further
development. Moreover, \citet{Korkut2015} began the development of a Turkish NLP
library named ``Guguk'', implemented in Haskell. I intend to merge this project
to Guguk, hence contributing towards the existence of a Haskell Turkish NLP
library; I also plan to continue working on this project as a part of Guguk.

Before I proceed, let me make a few remarks on the notational conventions I
follow throughout this text: I use \texttt{HMM} (with the fixed-width font) to
refer to the datatype I'm using from the \texttt{Data.HMM} package and HMM (with
regular font) to refer to the concept.

@@

% Briefly survey the most salient prior work that relates to the problem you
% wish to solve. This section should cite relevant sources. This section should
% be at least one to two paragraphs in length.
\cite{dincer2008suffix} built a POS tagger for Turkish using HMMs on suffixes.
This proves an effective approach seeing that (1) Turkish morphology is
exclusively based on suffixes, in contrast to to prefixes (2) morphology is
central to the Turkish language so it is likely that any given word is formed
through the agglutination of at least severals morphemes. The tagger described
by \citet{dincer2008suffix} has ``\%90.2 accuracy in the best case''.

@@

The project consists of two executables: (1) \texttt{create-model} that reads
the data from the corpus and parses it into a form from which one can easily
construct an \texttt{HMM}. Unfortunately, it is highly impractical---if not
impossible---to directly serialize the \texttt{HMM} datatype since it has three
fields containing functions.\footnote{See
  http://stackoverflow.com/questions/17785916/can-haskell-functions
  -be-serialized for further information.} I consider this a fault in the
design of the \texttt{Data.HMM} package and urge further development; this has
almost no effect on performance, but it obfuscates the code.

Since processing the corpus is a resource-demanding operation,
\texttt{create-model} is intended to be run only once after which
\texttt{run-tag} can be used to make use of the serialized components of the
HMM.

I should note that it is not possible to give a complete account of the
program's logic in this chapter although I explain the most salient components.
The source code is in the \texttt{ayberkt-final\_project} repository and is
well-commented. Should the reader have any concerns regarding the specifics of
the implementation they could find it convenient to consult the source code.

@@

The function \texttt{freqMap} found in \texttt{create/Main.hs} takes as input a
list of orderable elements (denoted with \texttt{Ord a}) and creates a
\texttt{Map} containing, as value, the number of occurrences of each element.

\begin{lstlisting}[
  language=Haskell,
  mathescape,
  columns=fullflexible,
  basicstyle=\small\ttfamily
]
freqMap :: Ord a $\Rightarrow$ [a] $\rightarrow$ Map a Int
freqMap elems = populate empty elems
  where populate :: Ord a Map a Int $\rightarrow$ [a] $\rightarrow$ Map a Int
         populate m [] = m
         populate m (x:xs) = let freq = (findWithDefault 0 x m) :: Int
                              in populate (insert x (freq + 1) m) xs
\end{lstlisting}

This function is called several times with lists of values designated to be
\emph{counted} (such as the list of tag bigrams, or the list of all tagged words)
et cetera. Pre-computing the counts is almost mandatory since frequencies (i.e.,
counts) will need to be accessed many times by the viterbi algorithm when
finding the most probable path through the model. Once the necessary maps are
generated, they are written to files in the directory \texttt{model}.

@@

\texttt{probability} when combined with \texttt{freqMap} forms an efficient
method for computing dependent probability.
